---
title: "Homework: Predicting Marine Fish Life-History Strategies with a Multilayer Perceptron"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

# Overview

Marine fish communities differ in their life-history strategies. Some communities are dominated by **fast-living species** that tend to grow quickly and mature earlier, while others are dominated by **slow-living species** that tend to mature later and live longer.

In this homework, you will use environmental conditions to predict whether a marine fish community is classified as:

- **slow-life-history community**
- **fast-life-history community**

You will build a **multilayer perceptron (MLP)** using R `torch` and compare it with a logistic-regression baseline.

## Learning objectives

By the end of this assignment, you should be able to:

1. Prepare tabular environmental data for a neural network.
2. Explain why feature scaling matters for MLPs.
3. Build and train an MLP for binary classification.
4. Use a validation set rather than the test set to make modeling decisions.
5. Diagnose underfitting and overfitting from training history.
6. Apply early stopping and regularization.
7. Compare an MLP with a simpler linear model.
8. Interpret model performance in an ecological context.

# Dataset

The file provided with this assignment is:

`FishTraits_MLP_Student.csv`

It contains the following columns:

| Variable | Description |
|---|---|
| `site_id` | Unique sampling/grid-cell identifier |
| `longitude`, `latitude` | Site coordinates |
| `Depth` | Water depth |
| `SBT` | Mean sea-bottom temperature |
| `SBS` | Mean sea-bottom salinity |
| `Chl` | Mean chlorophyll-a concentration |
| `SBT_var` | Seasonal variability in bottom temperature |
| `Chl_var` | Seasonal variability in chlorophyll-a |
| `Fishing` | Fishing-pressure indicator |
| `life_history_class` | Response class: `slow` or `fast` |
| `split` | Instructor-provided data split: `train`, `validation`, or `test` |
| `spatial_block` | Spatial-block identifier reserved for later spatial-validation exercises |

Use only these seven environmental predictors:

```text
Depth
SBT
SBS
Chl
SBT_var
Chl_var
Fishing
```

Do **not** use `site_id`, `longitude`, `latitude`, `split`, or `spatial_block` as predictors.

For modeling, encode:

- `slow` = 0
- `fast` = 1

The dataset already contains instructor-defined training, validation, and test sets. Use the `split` column exactly as provided.

# Submission

Submit your completed R Markdown document with:

- all code chunks executed,
- all requested figures,
- written answers to every question,
- and your final model comparison.

Your written answers should explain **what the model is doing and what the results mean**, not simply report numbers.

# 1. Load packages and confirm `torch` works

This assignment uses R `torch`.

If `torch` has not already been installed, run this once in the Console:

```text
install.packages("torch")
```

Complete any LibTorch/Lantern installation prompts, restart RStudio, and then run the code below.

```{r}
library(torch)
library(tidyverse)
x_check <- torch_tensor(
  c(1, 2, 3),
  dtype = torch_float32()
)

print(x_check)
```

If this prints a tensor, continue.

# 2. Load and inspect the data

```{r}
data_url <- "https://raw.githubusercontent.com/dblaskey/ML_Course_Code/main/Data/FishTraits.csv"

fish <- read.csv(
  data_url,
  stringsAsFactors = FALSE
)

cat("Rows:", nrow(fish), "\n")
cat("Columns:", ncol(fish), "\n")

head(fish)
```

```{r}
str(fish)
```

```{r}
table(fish$life_history_class)
```

```{r}
prop.table(
  table(fish$life_history_class)
)
```

### Question 1

1. How many observations are in the dataset?
2. Is the target approximately balanced?

# 3. Explore environmental differences between classes

```{r}
predictors <- c(
  "Depth",
  "SBT",
  "SBS",
  "Chl",
  "SBT_var",
  "Chl_var",
  "Fishing"
)

aggregate(
  fish[, predictors],
  by = list(
    life_history_class = fish$life_history_class
  ),
  FUN = mean,
  na.rm = TRUE
)
```

```{r}
variables_to_plot <- c(
  "SBT",
  "Depth",
  "SBT_var",
  "Chl"
)

par(
  mfrow = c(2, 2),
  mar = c(4, 4, 3, 1)
)

for (variable in variables_to_plot) {

  slow_values <- fish[
    fish$life_history_class == "slow",
    variable
  ]

  fast_values <- fish[
    fish$life_history_class == "fast",
    variable
  ]

  hist(
    slow_values,
    breaks = 25,
    probability = TRUE,
    main = variable,
    xlab = variable,
    col = rgb(0, 0, 0, 0.25),
    border = NA
  )

  hist(
    fast_values,
    breaks = 25,
    probability = TRUE,
    add = TRUE,
    col = rgb(0.6, 0.6, 0.6, 0.35),
    border = NA
  )

  legend(
    "topright",
    legend = c("slow", "fast"),
    fill = c(
      rgb(0, 0, 0, 0.25),
      rgb(0.6, 0.6, 0.6, 0.35)
    ),
    border = NA,
    bty = "n"
  )
}

par(mfrow = c(1, 1))
```

### Question 2

Describe at least **two environmental variables** that appear to differ between fast- and slow-life-history communities.

Do these patterns prove that the environmental variable *causes* the life-history strategy? Explain.

# 4. Define predictors and target

```{r}
X <- as.matrix(
  fish[, predictors]
)

y <- ifelse(
  fish$life_history_class == "fast",
  1,
  0
)

dim(X)
table(y)
```

# 5. Use the instructor-provided training, validation, and test sets

Use:

- `train` for model fitting,
- `validation` for model selection and early stopping,
- `test` for final evaluation only.

The test set must remain untouched while you compare model settings.

```{r}
train_mask <- fish$split == "train"
validation_mask <- fish$split == "validation"
test_mask <- fish$split == "test"

X_fit <- X[
  train_mask,
  ,
  drop = FALSE
]

X_validation <- X[
  validation_mask,
  ,
  drop = FALSE
]

X_test <- X[
  test_mask,
  ,
  drop = FALSE
]

y_fit <- y[train_mask]
y_validation <- y[validation_mask]
y_test <- y[test_mask]

cat("Fit:", nrow(X_fit), "\n")
cat("Validation:", nrow(X_validation), "\n")
cat("Test:", nrow(X_test), "\n")

table(
  fish$split,
  fish$life_history_class
)
```

# 6. Scale the environmental predictors

Fit the scaling parameters using the **training data only**.

```{r}
fit_means <- colMeans(X_fit)
fit_sds <- apply(
  X_fit,
  2,
  sd
)

X_fit_scaled <- sweep(
  sweep(
    X_fit,
    2,
    fit_means,
    "-"
  ),
  2,
  fit_sds,
  "/"
)

X_validation_scaled <- sweep(
  sweep(
    X_validation,
    2,
    fit_means,
    "-"
  ),
  2,
  fit_sds,
  "/"
)

X_test_scaled <- sweep(
  sweep(
    X_test,
    2,
    fit_means,
    "-"
  ),
  2,
  fit_sds,
  "/"
)
```

# 7. Establish a logistic-regression baseline

```{r}
logistic_fit_data <- data.frame(
  y = y_fit,
  X_fit_scaled
)

logistic_model <- glm(
  y ~ .,
  data = logistic_fit_data,
  family = binomial()
)

logistic_validation_probability <- predict(
  logistic_model,
  newdata = data.frame(
    X_validation_scaled
  ),
  type = "response"
)

logistic_validation_pred <- ifelse(
  logistic_validation_probability >= 0.5,
  1,
  0
)
```

We will use the following metric functions throughout the homework.

```{r}
accuracy_metric <- function(
  truth,
  prediction
) {
  mean(
    truth == prediction
  )
}

balanced_accuracy_metric <- function(
  truth,
  prediction
) {

  sensitivity <- mean(
    prediction[truth == 1] == 1
  )

  specificity <- mean(
    prediction[truth == 0] == 0
  )

  mean(
    c(
      sensitivity,
      specificity
    )
  )
}

roc_auc_metric <- function(
  truth,
  probability
) {

  n_positive <- sum(
    truth == 1
  )

  n_negative <- sum(
    truth == 0
  )

  probability_ranks <- rank(
    probability,
    ties.method = "average"
  )

  (
    sum(
      probability_ranks[
        truth == 1
      ]
    ) -
      n_positive *
        (n_positive + 1) /
        2
  ) /
    (
      n_positive *
        n_negative
    )
}
```

```{r}
cat(
  "Validation accuracy:",
  accuracy_metric(
    y_validation,
    logistic_validation_pred
  ),
  "\n"
)

cat(
  "Validation balanced accuracy:",
  balanced_accuracy_metric(
    y_validation,
    logistic_validation_pred
  ),
  "\n"
)

cat(
  "Validation ROC AUC:",
  roc_auc_metric(
    y_validation,
    logistic_validation_probability
  ),
  "\n"
)
```

### Question 3

Why is logistic regression a useful baseline before fitting an MLP?

What kind of decision relationship can an MLP represent that logistic regression cannot represent without manually creating nonlinear features?

# 8. Helper functions for the neural network

The R `torch` model will output **logits** rather than probabilities. During training, we will use `nn_bce_with_logits_loss()`, which combines the sigmoid transformation and binary cross-entropy loss in a numerically stable way.

```{r}
predict_torch_probability <- function(
  model,
  X
) {

  model$eval()

  X_tensor <- torch_tensor(
    X,
    dtype = torch_float32()
  )

  probability <- with_no_grad({

    logits <- model(
      X_tensor
    )

    torch_sigmoid(
      logits
    )
  })

  as.numeric(
    as_array(
      probability
    )
  )
}
```

```{r}
binary_cross_entropy <- function(
  truth,
  probability
) {

  epsilon <- 1e-7

  probability <- pmin(
    pmax(
      probability,
      epsilon
    ),
    1 - epsilon
  )

  -mean(
    truth *
      log(probability) +
      (1 - truth) *
      log(
        1 - probability
      )
  )
}
```

The training function below is supplied for you. It handles:

- Adam optimization,
- binary cross-entropy with logits,
- training and validation metrics,
- early stopping,
- `min_delta`,
- patience,
- and restoring the model from the epoch with the best validation loss.

```{r}
train_torch_model <- function(
  model,
  X_fit,
  y_fit,
  X_validation = NULL,
  y_validation = NULL,
  epochs = 200,
  learning_rate = 0.001,
  weight_decay = 0,
  early_stopping = FALSE,
  patience = 20,
  min_delta = 1e-4,
  restore_best_weights = TRUE,
  verbose = FALSE
) {

  optimizer <- optim_adam(
    model$parameters(),
    lr = learning_rate,
    weight_decay = weight_decay
  )

  loss_fn <- nn_bce_with_logits_loss()

  X_fit_tensor <- torch_tensor(
    X_fit,
    dtype = torch_float32()
  )

  y_fit_tensor <- torch_tensor(
    matrix(
      y_fit,
      ncol = 1
    ),
    dtype = torch_float32()
  )

  history <- data.frame(
    epoch = integer(),
    loss = numeric(),
    accuracy = numeric(),
    val_loss = numeric(),
    val_accuracy = numeric()
  )

  best_val_loss <- Inf
  best_epoch <- NA_integer_
  wait <- 0L

  best_model_path <- tempfile(
    fileext = ".pt"
  )

  for (epoch in seq_len(epochs)) {

    model$train()

    optimizer$zero_grad()

    logits <- model(
      X_fit_tensor
    )

    loss <- loss_fn(
      logits,
      y_fit_tensor
    )

    loss$backward()

    optimizer$step()

    train_probability <- predict_torch_probability(
      model,
      X_fit
    )

    train_prediction <- ifelse(
      train_probability >= 0.5,
      1,
      0
    )

    train_loss_value <- binary_cross_entropy(
      y_fit,
      train_probability
    )

    train_accuracy_value <- accuracy_metric(
      y_fit,
      train_prediction
    )

    if (!is.null(X_validation)) {

      validation_probability <- predict_torch_probability(
        model,
        X_validation
      )

      validation_prediction <- ifelse(
        validation_probability >= 0.5,
        1,
        0
      )

      validation_loss_value <- binary_cross_entropy(
        y_validation,
        validation_probability
      )

      validation_accuracy_value <- accuracy_metric(
        y_validation,
        validation_prediction
      )

    } else {

      validation_loss_value <- NA_real_
      validation_accuracy_value <- NA_real_
    }

    history <- rbind(
      history,
      data.frame(
        epoch = epoch,
        loss = train_loss_value,
        accuracy = train_accuracy_value,
        val_loss = validation_loss_value,
        val_accuracy = validation_accuracy_value
      )
    )

    if (
      early_stopping &&
      !is.null(X_validation)
    ) {

      if (
        validation_loss_value <
          best_val_loss -
          min_delta
      ) {

        best_val_loss <- validation_loss_value
        best_epoch <- epoch
        wait <- 0L

        torch_save(
          model,
          best_model_path
        )

      } else {

        wait <- wait + 1L
      }

      if (
        wait >= patience
      ) {

        if (verbose) {
          cat(
            "Early stopping at epoch",
            epoch,
            "\n"
          )
        }

        break
      }
    }

    if (
      verbose &&
      epoch %% 50 == 0
    ) {

      cat(
        "Epoch:",
        epoch,
        "validation loss:",
        round(
          validation_loss_value,
          4
        ),
        "\n"
      )
    }
  }

  stopped_epoch <- max(
    history$epoch
  )

  if (
    early_stopping &&
    restore_best_weights &&
    file.exists(
      best_model_path
    )
  ) {

    model <- torch_load(
      best_model_path
    )
  }

  if (
    file.exists(
      best_model_path
    )
  ) {

    unlink(
      best_model_path
    )
  }

  list(
    model = model,
    history = history,
    best_epoch = best_epoch,
    stopped_epoch = stopped_epoch
  )
}
```

# 9. Build your first MLP

Build a network with:

- 7 environmental inputs,
- one hidden layer with **8 units**,
- ReLU activation in the hidden layer,
- one output unit,
- Adam optimization,
- binary cross-entropy loss.

Remember: the output layer should return a **logit**. The sigmoid transformation is handled by the loss function during training and by `predict_torch_probability()` during prediction.

Complete the missing lines.

```{r}
mlp_net <- nn_module(
  "MLPNet",

  initialize = function(
    hidden_units = 8,
    dropout_rate = 0
  ) {

    # TODO: create the hidden layer
    self$hidden <- ______________________________

    self$dropout <- nn_dropout(
      p = dropout_rate
    )

    # TODO: create the one-unit output layer
    self$output <- ______________________________
  },

  forward = function(x) {

    # TODO: apply the hidden layer
    x <- ______________________________

    # TODO: apply ReLU
    x <- ______________________________

    x <- self$dropout(
      x
    )

    self$output(
      x
    )
  }
)
```

```{r}
set.seed(42)
torch_manual_seed(42)

model <- mlp_net(
  hidden_units = 8,
  dropout_rate = 0
)

model
```

# 10. Train the MLP

```{r}
fitted_model <- train_torch_model(
  model = model,
  X_fit = X_fit_scaled,
  y_fit = y_fit,
  X_validation = X_validation_scaled,
  y_validation = y_validation,
  epochs = 200,
  learning_rate = 0.01
)

history_df <- fitted_model$history
model <- fitted_model$model
```

# 11. Visualize the learning history

```{r}
par(
  mfrow = c(1, 2)
)

plot(
  history_df$epoch,
  history_df$loss,
  type = "l",
  xlab = "Epoch",
  ylab = "Binary cross-entropy loss",
  main = "Loss"
)

lines(
  history_df$epoch,
  history_df$val_loss,
  lty = 2
)

legend(
  "topright",
  legend = c(
    "training",
    "validation"
  ),
  lty = c(
    1,
    2
  ),
  bty = "n"
)

plot(
  history_df$epoch,
  history_df$accuracy,
  type = "l",
  xlab = "Epoch",
  ylab = "Accuracy",
  main = "Accuracy"
)

lines(
  history_df$epoch,
  history_df$val_accuracy,
  lty = 2
)

legend(
  "bottomright",
  legend = c(
    "training",
    "validation"
  ),
  lty = c(
    1,
    2
  ),
  bty = "n"
)

par(
  mfrow = c(1, 1)
)
```

### Question 4

Examine both plots.

1. Does the model appear to be underfitting, overfitting, or reasonably fit?
2. Around what epoch does validation loss stop improving substantially?
3. Why can validation accuracy remain nearly constant while validation loss continues to change?

# 12. Add early stopping

Now train the same model with early stopping.

Monitor **validation loss** and use:

- `patience = 20`
- `min_delta = 1e-4`
- `restore_best_weights = TRUE`
- maximum of 1000 epochs

Fill in the missing values.

```{r}
set.seed(42)
torch_manual_seed(42)

early_model <- mlp_net(
  hidden_units = 8,
  dropout_rate = 0
)

early_fitted <- train_torch_model(
  model = early_model,
  X_fit = X_fit_scaled,
  y_fit = y_fit,
  X_validation = X_validation_scaled,
  y_validation = y_validation,
  epochs = ______,                # TODO
  learning_rate = 0.01,
  early_stopping = TRUE,
  patience = ______,              # TODO
  min_delta = ______,             # TODO
  restore_best_weights = ______   # TODO
)

early_history_df <- early_fitted$history
early_model <- early_fitted$model
```

```{r}
best_epoch <- early_fitted$best_epoch
stopped_epoch <- early_fitted$stopped_epoch

plot(
  early_history_df$epoch,
  early_history_df$val_loss,
  type = "l",
  xlab = "Epoch",
  ylab = "Validation loss",
  main = "Early stopping based on validation loss"
)

abline(
  v = best_epoch,
  lty = 2
)

abline(
  v = stopped_epoch,
  lty = 3
)

legend(
  "topright",
  legend = c(
    "validation loss",
    "best validation loss",
    "training stopped"
  ),
  lty = c(
    1,
    2,
    3
  ),
  bty = "n"
)

cat(
  "Best epoch:",
  best_epoch,
  "\n"
)

cat(
  "Stopped epoch:",
  stopped_epoch,
  "\n"
)
```

### Question 5

1. At what epoch was validation loss lowest?
2. At what epoch did training stop?
3. Why are those two epochs not necessarily identical?
4. What does `restore_best_weights = TRUE` do?

# 13. Experiment with model complexity

The helper below builds the same MLP with different hidden-unit, dropout, and L2 settings.

In R `torch`, L2 regularization is supplied to Adam as `weight_decay`.

```{r}
build_mlp <- function(
  hidden_units = 8,
  dropout_rate = 0
) {

  mlp_net(
    hidden_units = hidden_units,
    dropout_rate = dropout_rate
  )
}
```

```{r}
evaluate_setting <- function(
  label,
  hidden_units = 8,
  dropout_rate = 0,
  l2_strength = 0
) {

  set.seed(42)
  torch_manual_seed(42)

  model <- build_mlp(
    hidden_units = hidden_units,
    dropout_rate = dropout_rate
  )

  fitted <- train_torch_model(
    model = model,
    X_fit = X_fit_scaled,
    y_fit = y_fit,
    X_validation = X_validation_scaled,
    y_validation = y_validation,
    epochs = 1000,
    learning_rate = 0.01,
    weight_decay = l2_strength,
    early_stopping = TRUE,
    patience = 20,
    min_delta = 1e-4,
    restore_best_weights = TRUE
  )

  probability <- predict_torch_probability(
    fitted$model,
    X_validation_scaled
  )

  prediction <- ifelse(
    probability >= 0.5,
    1,
    0
  )

  data.frame(
    setting = label,
    hidden_units = hidden_units,
    dropout = dropout_rate,
    l2 = l2_strength,
    epochs_used = fitted$stopped_epoch,
    validation_accuracy = accuracy_metric(
      y_validation,
      prediction
    ),
    validation_balanced_accuracy = balanced_accuracy_metric(
      y_validation,
      prediction
    ),
    validation_auc = roc_auc_metric(
      y_validation,
      probability
    ),
    best_validation_loss = min(
      fitted$history$val_loss
    )
  )
}
```

Change **one setting at a time**. For each experiment, train using the same fit and validation data and record the validation results.

Complete at least these experiments:

1. `hidden_units = 1`
2. `hidden_units = 32`
3. `hidden_units = 64`
4. baseline 8-unit model + dropout of `0.30`
5. baseline 8-unit model + L2 regularization

Use early stopping for each experiment.

```{r}
experiments <- rbind(

  # TODO: add the required experimental settings here.
  #
  # Example:
  #
  # evaluate_setting(
  #   "1 hidden unit",
  #   hidden_units = 1
  # )
  # variables to change hidden_units, dropout_rate, and l2_strength

)

results <- experiments

results
```

### Question 6

Using only the **validation results**:

1. Does the 1-unit model appear to underfit?
2. Does increasing the number of hidden units consistently improve validation performance?
3. Did dropout help?
4. Did L2 regularization help?
5. Which model setting would you carry forward for final evaluation, and why?

# 14. Final model evaluation

Choose the model setting you selected using validation performance.

Refit that model using the same fit/validation workflow and early stopping. Then evaluate it **once** on the test set.

```{r}
# TODO: replace these values with the settings you selected.

FINAL_HIDDEN_UNITS <- ______
FINAL_DROPOUT <- ______
FINAL_L2 <- ______
```

```{r}
set.seed(42)
torch_manual_seed(42)

final_model <- build_mlp(
  hidden_units = FINAL_HIDDEN_UNITS,
  dropout_rate = FINAL_DROPOUT
)

final_fitted <- train_torch_model(
  model = final_model,
  X_fit = X_fit_scaled,
  y_fit = y_fit,
  X_validation = X_validation_scaled,
  y_validation = y_validation,
  epochs = 1000,
  learning_rate = 0.01,
  weight_decay = FINAL_L2,
  early_stopping = TRUE,
  patience = 20,
  min_delta = 1e-4,
  restore_best_weights = TRUE
)

final_model <- final_fitted$model

test_probability <- predict_torch_probability(
  final_model,
  X_test_scaled
)

test_pred <- ifelse(
  test_probability >= 0.5,
  1,
  0
)

cat(
  "Test accuracy:",
  accuracy_metric(
    y_test,
    test_pred
  ),
  "\n"
)

cat(
  "Test balanced accuracy:",
  balanced_accuracy_metric(
    y_test,
    test_pred
  ),
  "\n"
)

cat(
  "Test ROC AUC:",
  roc_auc_metric(
    y_test,
    test_probability
  ),
  "\n"
)
```

```{r}
confusion_matrix <- table(
  Predicted = factor(
    test_pred,
    levels = c(0, 1),
    labels = c(
      "slow",
      "fast"
    )
  ),
  Observed = factor(
    y_test,
    levels = c(0, 1),
    labels = c(
      "slow",
      "fast"
    )
  )
)

confusion_matrix
```

# 15. Compare the MLP with logistic regression on the test set

Only now should you evaluate the logistic-regression baseline on the test set.

```{r}
logistic_test_probability <- predict(
  logistic_model,
  newdata = data.frame(
    X_test_scaled
  ),
  type = "response"
)

logistic_test_pred <- ifelse(
  logistic_test_probability >= 0.5,
  1,
  0
)

comparison <- data.frame(
  model = c(
    "Logistic regression",
    "MLP"
  ),

  test_accuracy = c(
    accuracy_metric(
      y_test,
      logistic_test_pred
    ),
    accuracy_metric(
      y_test,
      test_pred
    )
  ),

  test_balanced_accuracy = c(
    balanced_accuracy_metric(
      y_test,
      logistic_test_pred
    ),
    balanced_accuracy_metric(
      y_test,
      test_pred
    )
  ),

  test_roc_auc = c(
    roc_auc_metric(
      y_test,
      logistic_test_probability
    ),
    roc_auc_metric(
      y_test,
      test_probability
    )
  )
)

comparison
```

### Question 7

Did the MLP clearly outperform logistic regression?

Your answer should consider:

- accuracy,
- balanced accuracy,
- ROC AUC,
- model complexity,
- and whether the improvement is large enough to justify the more complicated model.

A neural network is not automatically the better scientific model simply because it is more flexible.

# Grading guide

| Component | Points |
|---|---:|
| Data inspection | 1 |
| Train/validation/test design and scaling | 1 |
| Logistic-regression baseline | 1 |
| Correct MLP construction and training | 2 |
| Learning curves and early stopping | 4 |
| Model-complexity / regularization experiments | 4 |
| Final test evaluation and baseline comparison | 2 |
| Questions 1-7 | 10 |
| **Total** | **25** |
