---
title: "Homework 1"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r packages}
library(tidyverse)
library(lubridate)
```

## Warm up

Given the vector below:

```{r quiz-scores}
quiz_scores <- c(78, 85, 92, 67, 88, 95, 73, 84)
```

Write R code to:

1.  Print the first score.
2.  Print the last score.
3.  Calculate the average score.
4.  Create a new vector containing only scores greater than or equal to 80.
5.  Count how many students scored below 80.

```{r warm-up-answers}

```

Write a function called `letter_grade()` that takes a numeric score and returns:

| Score Range | Grade |
|:------------|:------|
| 90 or above | A     |
| 80–89       | B     |
| 70–79       | C     |
| 60–69       | D     |
| Below 60    | F     |

Example:

``` r
letter_grade(92)
```

Expected output:

``` r
[1] "A"
```

```{r letter-grade-function}

```

Then:

7.  Create a data frame using the `quiz_scores` vector.
8.  Create a new column named `Letter_Grade` by applying the `letter_grade()` function.

```{r quiz-data-frame}

```

Add the following vector as a column named `HoursOfStudy`:

```{r study-hours}
study_hours <- c("1", "1.5", "2.5", "0", "1", "1.5", "0.5", "1")
```

Once you create the column, do the following:

9.  Change the column to numeric.
10. Create a plot using `HoursOfStudy` and `quiz_scores`.
11. Make the plot look polished and readable.

```{r study-hours-plot}

```

## Data Analysis

For this part, we will work with an actual dataset.

Load the following dataset:

<https://raw.githubusercontent.com/dblaskey/ML_Course_Code/main/Data/weatherHistory.csv>

```{r load-weather-data}

```

It is usually best practice to keep a raw copy of the original data unless the dataset is exceptionally large. Create a copy of the dataset.

```{r copy-weather-data}

```

Clean the column names of the data frame:

1.  Remove spaces at the beginning and end of each column name.
2.  Convert all column names to lowercase.
3.  Replace any group of characters that is not a letter or number with an underscore.
4.  Remove underscores from the beginning and end.

```{r clean-column-names}

```

Now that you have cleaned the column names, it is time to understand and clean the dataset:

1.  Notice there seems to be a typo in a column heading. Change `loud_cover` to `cloud_cover`.
2.  Check for impossible data, such as precipitation, wind speed, visibility, cloud cover, or pressure below 0, and any variable outside reasonable earthly maximum and minimum values.
3.  Give the count, mean, standard deviation, maximum, and minimum values for each numeric column.
4.  Count the number of missing values in each column.
5.  Create a new column that converts `formatted_date` to a date-time value.
6.  Create new columns for hour, day, month, and year.

```{r clean-weather-data}

```

Now create multiple plots to continue exploring the dataset:

1.  Plot a time-series graph of wind speed.
2.  Create a second time-series graph of air temperature and humidity using two different axes.
3.  Plot daily average temperature by calendar date; for example, average all January 1 observations, all January 2 observations, and so on.
4.  Make the plots polished and readable.

```{r weather-plots}

```

## Rubric

| Component             | Points |
|:----------------------|-------:|
| Warm up               |      5 |
| Load and copy data    |      2 |
| Clean column names    |      3 |
| Clean dataset         |      6 |
| Plots (3 points each) |      9 |
| **Total**             | **25** |
