# Lecture 1 - In Class Activity

**Goal:** Build a small, reproducible R project directory and Git repository.

By the end of this document, you will have created a project folder with:

``` text
GEOL599_ML/
├── data/
│   └── raw/
├── notebooks/
├── outputs/
├── scripts/
├── .gitignore
└── README.md
```

> **Important:** This document uses R code chunks. Terminal commands are run from R with `system2()`. For those in RStudio, make sure to switch from `Source` to `Visual` to see this markdown document how it is supposed to me loaded. Also, double check that `Knit on Save` is left unchecked.

## Step 0: Check your setup

Run the chunks below to confirm that R and Git are available.

If Git is not installed, tell your instructor before continuing.

```{r check-r}
R.version.string
```

```{r check-git}
git_version <- system2("git", "--version", stdout = TRUE, stderr = TRUE)
cat(git_version, sep = "\n")
```

If you have an old version of R or do not have Git, please update or download it now.\
Let Dylan know if you have any problems.

## Step 1: Create a project directory

We will create a new folder called `GEOL599_ML`.

In a normal terminal, you would run:

``` bash
mkdir GEOL599_ML
cd GEOL599_ML
```

Inside R Markdown, we can create the folder with R and set it as the root directory for the remaining code chunks.

```{r create-project-directory}
project_dir <- file.path(getwd(), "GEOL599_ML")
dir.create(project_dir, showWarnings = FALSE, recursive = TRUE)
normalizePath(project_dir)

knitr::opts_knit$set(root.dir = project_dir)
```

## Step 2: Create a clean project structure

A consistent folder structure makes projects easier to understand, rerun, share, and debug.

We will create four main directories:

- `data/raw/` for original input data
- `notebooks/` for exploratory notebooks
- `outputs/` for generated files, figures, and results
- `scripts/` for reusable R scripts

```{r create-folders}
folders <- c("data/raw", "notebooks", "outputs", "scripts")

for (folder in folders) {
  dir.create(folder, recursive = TRUE, showWarnings = FALSE)
}

project_folders <- list.dirs(".", recursive = TRUE, full.names = TRUE)
cat(sort(project_folders), sep = "\n")
```

### Checkpoint 1

You should now have a folder structure that looks like this:

``` text
GEOL599_ML/
├── data/
│   └── raw/
├── notebooks/
├── outputs/
└── scripts/
```

**Reflection prompt:** Why might it be risky to keep scripts, datasets, outputs, and notes all in one folder?

## Step 3: Create a README file

A `README.md` file explains what the project is, how to use it, and what someone needs to know to reproduce your work.

```{r create-readme}
readme_text <- c(
  "# GEOL599 Applied Machine Learning for Geoscience Data",
  "",
  "This project includes all in-class assignments and homework from this course.",
  "",
  "## Project structure",
  "",
  "- `data/raw/`: original input data",
  "- `notebooks/`: exploratory notebooks",
  "- `outputs/`: generated results and figures",
  "- `scripts/`: reusable R scripts"
)

writeLines(readme_text, "README.md")
cat(readLines("README.md"), sep = "\n")
```

### Reflection prompt

What information would someone need to understand or rerun your project six months from now?

## Step 4: Initialize Git

Git tracks changes to your files over time.

First, initialize a Git repository.

```{r git-init}
cat(system2("git", "init", stdout = TRUE, stderr = TRUE), sep = "\n")
```

```{r git-status-initial}
cat(system2("git", "status", stdout = TRUE, stderr = TRUE), sep = "\n")
```

## Step 5: Create a `.gitignore` file

Not every file belongs in Git.

For example, we usually do **not** commit:

- R project environment files
- R history and workspace files
- R Markdown cache files
- Generated outputs
- System files such as `.DS_Store`

```{r create-gitignore}
gitignore_text <- c(
  ".Rproj.user/",
  ".Rhistory",
  ".RData",
  ".Ruserdata",
  "*_cache/",
  "*_files/",
  "outputs/",
  ".DS_Store"
)

writeLines(gitignore_text, ".gitignore")
cat(readLines(".gitignore"), sep = "\n")
```

```{r git-status-after-ignore}
cat(system2("git", "status", stdout = TRUE, stderr = TRUE), sep = "\n")
```

## Step 6: Make the first commit

A commit is a snapshot of your project at a meaningful point in time.

Before committing, Git may require your name and email. If Git asks for these, run the following commands in a terminal, replacing the values with your own:

``` bash
git config --global user.name "Your Name"
git config --global user.email "your.email@example.com"
```

Now add the important project files and commit them.

```{r git-add}
cat(
  system2("git", c("add", "README.md", ".gitignore"), stdout = TRUE, stderr = TRUE),
  sep = "\n"
)
cat(system2("git", "status", stdout = TRUE, stderr = TRUE), sep = "\n")
```

```{r git-commit}
commit_result <- system2(
  "git",
  c("commit", "-m", shQuote("Initial project setup")),
  stdout = TRUE,
  stderr = TRUE
)
cat(commit_result, sep = "\n")
```

```{r git-log}
cat(system2("git", c("log", "--oneline"), stdout = TRUE, stderr = TRUE), sep = "\n")
```

## Optional: Connect to GitHub

If you have a GitHub account and have created an empty repository online, you can connect this local repository to GitHub.

In a terminal, run:

``` bash
git branch -M main
git remote add origin <repository-url>
git push -u origin main
```

Replace `<repository-url>` with your actual GitHub repository URL.

This step is optional because account setup and authentication can take extra time. However, you should set this up before the next class.
