---
title: "In-Class Activity 6: Decision Trees, Random Forests, and Boosting"
author: "GEOL 599: Machine Learning"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
# If needed, install these packages first:
# install.packages(c("tidyverse", "tidymodels", "rpart", "rpart.plot", "ranger", "xgboost", "patchwork"))

library(tidyverse)
library(tidymodels)
library(rpart)
library(rpart.plot)
library(ranger)
library(xgboost)
library(patchwork)

tidymodels_prefer()
set.seed(42)
```

# Overview

In this activity, you will use the **Statlog Landsat Satellite** dataset to compare several tree-based machine-learning models:

- A single decision tree
- Random forest
- XGBoost

Each row represents a small 3 by 3 pixel neighborhood from Landsat satellite imagery. The predictors are spectral measurements from those pixels, and the target is the land-cover class of the central pixel.

## Learning Goals

By the end of this activity, you should be able to:

- Explain why decision trees make **axis-aligned splits**.
- Visualize the limitations of a single decision tree boundary.
- Compare a single tree with ensemble tree methods.
- Explain the difference between bagging and boosting.
- Use train/test splits and cross-validation appropriately.
- Tune tree-based models without leaking information from the test set.
- Interpret model behavior using feature importance and permutation importance.

# 1. Load the Landsat Dataset

The dataset is the Statlog Landsat Satellite dataset from OpenML/UCI. Each row contains 36 spectral measurements:

```text
9 pixels x 4 spectral bands = 36 predictors
```

The four spectral values for the central pixel are attributes 17, 18, 19, and 20. With the naming scheme below, those are:

```text
pixel_5_band_1
pixel_5_band_2
pixel_5_band_3
pixel_5_band_4
```

```{r load-data}
data_url <- "https://raw.githubusercontent.com/dblaskey/ML_Course_Code/main/Data/Landsat_Landcover.csv"

landsat_raw <- read_csv(data_url, show_col_types = FALSE)

glimpse(landsat_raw)
```

```{r clean-data}
feature_cols <- paste0(
  "pixel_",
  rep(1:9, each = 4),
  "_band_",
  rep(1:4, times = 9)
)

# The target is the last column in this CSV.
target_col <- names(landsat_raw)[ncol(landsat_raw)]

landsat <- landsat_raw %>%
  rename_with(~ feature_cols, .cols = 1:36) %>%
  rename(land_cover_code = all_of(target_col)) %>%
  mutate(
    land_cover_raw = as.character(land_cover_code),
    land_cover_clean = land_cover_raw %>%
      str_to_lower() %>%
      str_replace_all("[._-]+", " ") %>%
      str_squish(),
    land_cover_number = readr::parse_number(land_cover_raw),
    land_cover = case_when(
      land_cover_number == 1~ "red_soil",
      land_cover_number == 2 ~ "cotton_crop",
      land_cover_number == 3 ~ "grey_soil",
      land_cover_number == 4 ~ "damp_grey_soil",
      land_cover_number == 5 ~ "soil_with_vegetation_stubble",
      land_cover_number == 7 ~ "very_damp_grey_soil",
      TRUE ~ NA_character_
    ),
    land_cover = factor(land_cover)
  ) %>%
  filter(!is.na(land_cover)) %>%
  select(all_of(feature_cols), land_cover)

landsat %>% slice_head(n = 6)
```

# 2. Initial Data Checks

Before modeling, check the number of rows, columns, missing values, and class balance.

```{r data-checks}
cat("Rows and columns:\n")
dim(landsat)

cat("\nMissing values:\n")
sum(is.na(landsat))

class_counts <- landsat %>%
  count(land_cover, sort = TRUE)

class_counts
```

```{r class-count-plot, fig.width=8, fig.height=4}
ggplot(class_counts, aes(x = reorder(land_cover, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Landsat land-cover class counts",
    x = "Land-cover class",
    y = "Number of observations"
  ) +
  theme_minimal(base_size = 13)
```

## Question 1

Is this classification problem balanced? Which land-cover classes are most and least common?

# 3. Why Decision Trees Have Axis-Aligned Splits

Decision trees split one feature at a time.

Examples:

```text
pixel_5_band_1 <= 66
pixel_5_band_2 <= 42
pixel_5_band_4 <= 87
```

That means a decision tree creates boundaries that are vertical or horizontal in a two-feature plot. These are called **axis-aligned splits**.

This is powerful and interpretable, but it can be inefficient when the real class boundary is diagonal, curved, or depends on a smooth combination of predictors.

# 4. Build a Two-Feature Dataset for Visualization

To visualize decision boundaries, we will use only two predictors from the central pixel:

- `pixel_5_band_1`
- `pixel_5_band_2`

This is intentionally limited. Later, we will use all 36 spectral predictors.

```{r two-feature-data}
two_feature_cols <- c("pixel_5_band_1", "pixel_5_band_2")

landsat_two <- landsat %>%
  select(all_of(two_feature_cols), land_cover) %>%
  drop_na()

set.seed(42)
two_split <- initial_split(landsat_two, prop = 0.75, strata = land_cover)
two_train <- training(two_split)
two_test <- testing(two_split)

class_levels <- levels(landsat_two$land_cover)
class_palette <- setNames(scales::hue_pal()(length(class_levels)), class_levels)

class_palette
```

```{r two-feature-plot, fig.width=8, fig.height=6}
ggplot(landsat_two, aes(x = pixel_5_band_1, y = pixel_5_band_2, color = land_cover)) +
  geom_point(alpha = 0.65, size = 1.6) +
  scale_color_manual(values = class_palette) +
  labs(
    title = "Landsat classes using two central-pixel spectral bands",
    x = "pixel_5_band_1",
    y = "pixel_5_band_2",
    color = "Land cover"
  ) +
  theme_minimal(base_size = 13)
```

## Question 2

Looking at the scatterplot, where might a single decision tree struggle? Do the classes appear separable by only horizontal and vertical cuts?

# 5. Fit a Shallow Decision Tree

A shallow tree is easy to interpret, but it may be too simple.

```{r shallow-tree}
shallow_tree <- rpart(
  land_cover ~ pixel_5_band_1 + pixel_5_band_2,
  data = two_train,
  method = "class",
  control = rpart.control(maxdepth = 2, cp = 0.001)
)

shallow_tree_pred_train <- predict(shallow_tree, newdata = two_train, type = "class")
shallow_tree_pred_test <- predict(shallow_tree, newdata = two_test, type = "class")

mean(shallow_tree_pred_train == two_train$land_cover)
mean(shallow_tree_pred_test == two_test$land_cover)
```

```{r plot-shallow-tree, fig.width=14, fig.height=7}
rpart.plot(
  shallow_tree,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  main = "Shallow decision tree"
)
```

# 6. Decision Boundary Helper Function

The background color shows the model's predicted class across the two-feature space. The points show the observed classes. Training observations are circles and test observations are X symbols.

```{r decision-boundary-function}
predict_tree_classes <- function(model, new_data) {
  predict(model, newdata = new_data, type = "class")
}


plot_decision_boundary <- function(model, train_data, test_data, title, predict_fun = predict_tree_classes) {
  x_min <- min(train_data$pixel_5_band_1, test_data$pixel_5_band_1) - 5
  x_max <- max(train_data$pixel_5_band_1, test_data$pixel_5_band_1) + 5
  y_min <- min(train_data$pixel_5_band_2, test_data$pixel_5_band_2) - 5
  y_max <- max(train_data$pixel_5_band_2, test_data$pixel_5_band_2) + 5

  prediction_grid <- expand_grid(
    pixel_5_band_1 = seq(x_min, x_max, length.out = 300),
    pixel_5_band_2 = seq(y_min, y_max, length.out = 300)
  )

  grid_predictions <- prediction_grid %>%
    mutate(.pred_class = predict_fun(model, prediction_grid))

  point_data <- bind_rows(
    train_data %>% mutate(split = "Training"),
    test_data %>% mutate(split = "Test")
  )

  ggplot() +
    geom_raster(
      data = grid_predictions,
      aes(x = pixel_5_band_1, y = pixel_5_band_2, fill = .pred_class),
      alpha = 0.25
    ) +
    geom_point(
      data = point_data,
      aes(x = pixel_5_band_1, y = pixel_5_band_2, color = land_cover, shape = split),
      size = 1.8,
      alpha = 0.8
    ) +
    scale_fill_manual(values = class_palette, drop = FALSE) +
    scale_color_manual(values = class_palette, drop = FALSE) +
    scale_shape_manual(values = c("Training" = 16, "Test" = 4)) +
    coord_cartesian(xlim = c(-2.5, 2.5), ylim = c(-2.5, 2.5)) +
    labs(
      title = title,
      x = "pixel_5_band_1",
      y = "pixel_5_band_2",
      fill = "Predicted class",
      color = "Observed class",
      shape = "Dataset split"
    ) +
    theme_minimal(base_size = 12)
}
```

```{r shallow-tree-boundary, fig.width=8, fig.height=6}
plot_decision_boundary(
  shallow_tree,
  two_train,
  two_test,
  "Shallow decision tree: axis-aligned splits"
)
```

## Question 3

What do you notice about the decision boundary? Why does it look like rectangles or stair steps?

# 7. Compare Single Tree Depths

Deeper trees are more flexible, but they can overfit.

```{r tree-depths, fig.width=12, fig.height=10}
tree_depths <- c(1, 2, 5, 30)

tree_depth_plots <- map(
  tree_depths,
  function(depth) {
    tree_fit <- rpart(
      land_cover ~ pixel_5_band_1 + pixel_5_band_2,
      data = two_train,
      method = "class",
      control = rpart.control(maxdepth = depth, cp = 0.001)
    )

    plot_decision_boundary(
      tree_fit,
      two_train,
      two_test,
      paste0("Decision tree maxdepth = ", depth)
    )
  }
)

wrap_plots(tree_depth_plots, ncol = 2)
```

## Question 4

How does the decision boundary change as the tree gets deeper? Which model seems most likely to overfit?

# 8. Compare Tree-Based Ensembles on the Same Two Features

Now compare:

- Decision tree
- Random forest
- XGBoost

This section is still using only two features so the boundaries can be visualized.

```{r two-feature-models}
# Decision tree
boundary_tree <- rpart(
  land_cover ~ pixel_5_band_1 + pixel_5_band_2,
  data = two_train,
  method = "class",
  control = rpart.control(maxdepth = 5, cp = 0.001)
)

# Random forest
boundary_rf <- ranger(
  land_cover ~ pixel_5_band_1 + pixel_5_band_2,
  data = two_train,
  num.trees = 200,
  max.depth = 6,
  probability = FALSE,
  seed = 42
)


# XGBoost through tidymodels
boundary_xgb_spec <- boost_tree(
  trees = 100,
  tree_depth = 2,
  learn_rate = 0.1,
  sample_size = 0.9,
  mtry = 2
) %>%
  set_mode("classification") %>%
  set_engine("xgboost", eval_metric = "mlogloss")

boundary_xgb_wf <- workflow() %>%
  add_formula(land_cover ~ pixel_5_band_1 + pixel_5_band_2) %>%
  add_model(boundary_xgb_spec)

boundary_xgb <- fit(boundary_xgb_wf, data = two_train)
```

```{r two-feature-accuracy}
predict_ranger_classes <- function(model, new_data) {
  predict(model, data = new_data)$predictions %>%
    factor(levels = class_levels)
}

predict_workflow_classes <- function(model, new_data) {
  predict(model, new_data, type = "class")$.pred_class %>%
    factor(levels = class_levels)
}

two_feature_accuracy <- tibble(
  model = c("Decision tree", "Random forest", "XGBoost"),
  train_accuracy = c(
    mean(predict_tree_classes(boundary_tree, two_train) == two_train$land_cover),
    mean(predict_ranger_classes(boundary_rf, two_train) == two_train$land_cover),
    mean(predict_workflow_classes(boundary_xgb, two_train) == two_train$land_cover)
  ),
  test_accuracy = c(
    mean(predict_tree_classes(boundary_tree, two_test) == two_test$land_cover),
    mean(predict_ranger_classes(boundary_rf, two_test) == two_test$land_cover),
    mean(predict_workflow_classes(boundary_xgb, two_test) == two_test$land_cover)
  )
)

two_feature_accuracy
```

```{r two-feature-boundaries, fig.width=14, fig.height=11}
p1 <- plot_decision_boundary(boundary_tree, two_train, two_test, "Decision tree")
p2 <- plot_decision_boundary(boundary_rf, two_train, two_test, "Random forest", predict_fun = predict_ranger_classes)
p3 <- plot_decision_boundary(boundary_xgb, two_train, two_test, "XGBoost", predict_fun = predict_workflow_classes)

p1 / p2 / p3
```

## Question 5

Compare the three decision boundaries.

1. Which model has the simplest boundary?
2. Which model has the most complex boundary?
3. Which model seems most likely to overfit?
4. Why do all tree-based boundaries still look somewhat blocky?

# Part 2: Best Practices With All Spectral Features

The two-feature plots are useful for learning, but they are not the best possible models. In real modeling, we should use a proper train/test split, cross-validation, preprocessing, and tuning.

We will now use all 36 spectral predictors.

Best-practice reminders:

- Keep the test set untouched until final evaluation.
- Use cross-validation for model comparison and tuning.
- Use workflows or consistent preprocessing to avoid leakage.
- Tune complexity controls such as tree depth, number of trees, and learning rate.
- Compare against a simple baseline.

# 9. Prepare the Full Feature Dataset

```{r full-feature-data}
set.seed(42)
landsat_split <- initial_split(landsat, prop = 0.75, strata = land_cover)
train_data <- training(landsat_split)
test_data <- testing(landsat_split)

nrow(train_data)
nrow(test_data)
ncol(train_data) - 1
```

# 10. Define Baseline and Candidate Models

Tree models do not require feature scaling. We include a median imputer in the recipe as a best-practice safeguard, although this dataset should not have missing predictor values.

```{r model-workflows}
landsat_recipe <- recipe(land_cover ~ ., data = train_data) %>%
  step_impute_median(all_numeric_predictors())

baseline_spec <- null_model() %>%
  set_mode("classification") %>%
  set_engine("parsnip")

tree_spec <- decision_tree(
  tree_depth = 8,
  min_n = 5,
  cost_complexity = 0.001
) %>%
  set_mode("classification") %>%
  set_engine("rpart")

rf_spec <- rand_forest(
  trees = 300,
  mtry = 6,
  min_n = 2
) %>%
  set_mode("classification") %>%
  set_engine("ranger")

xgb_spec <- boost_tree(
  trees = 150,
  tree_depth = 2,
  learn_rate = 0.05,
  sample_size = 0.9,
  mtry = 24
) %>%
  set_mode("classification") %>%
  set_engine("xgboost", eval_metric = "mlogloss")

make_workflow <- function(model_spec) {
  workflow() %>%
    add_recipe(landsat_recipe) %>%
    add_model(model_spec)
}

candidate_workflows <- list(
  "Baseline" = make_workflow(baseline_spec),
  "Decision tree" = make_workflow(tree_spec),
  "Random forest" = make_workflow(rf_spec),
  "XGBoost" = make_workflow(xgb_spec)
)
```

# 11. Cross-Validation Comparison

Use cross-validation on the training set to compare models before touching the test set.

```{r cross-validation}
set.seed(42)
cv_folds <- vfold_cv(train_data, v = 5, strata = land_cover)

classification_metrics <- metric_set(accuracy, f_meas)

cv_results <- imap_dfr(
  candidate_workflows,
  function(wf, model_name) {
    fit_resamples(
      wf,
      resamples = cv_folds,
      metrics = classification_metrics,
      control = control_resamples(save_pred = TRUE)
    ) %>%
      collect_metrics() %>%
      mutate(model = model_name)
  }
)

cv_results
```

```{r cv-summary}
cv_summary <- cv_results %>%
  select(model, .metric, mean, std_err) %>%
  arrange(.metric, desc(mean))

cv_summary
```

## Question 6

Which model has the best cross-validated macro F1 score? Does any model have much higher training accuracy than validation accuracy? What would that suggest?

# 12. Tune XGBoost

XGBoost has many tuning parameters. We will tune only a few so the activity stays manageable.

Important parameters:

- `tree_depth`: tree complexity
- `learn_rate`: contribution of each tree
- `trees`: number of boosting rounds
- `sample_size`: fraction of rows used per tree
- `mtry`: number of predictors sampled at each split

Best practice: tune on the training set using cross-validation, then evaluate once on the held-out test set.

```{r tune-xgboost}
xgb_tune_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  sample_size = tune(),
  mtry = tune(),
  min_n = 2
) %>%
  set_mode("classification") %>%
  set_engine("xgboost", eval_metric = "mlogloss")

xgb_tune_wf <- workflow() %>%
  add_recipe(landsat_recipe) %>%
  add_model(xgb_tune_spec)

xgb_grid <- crossing(
  trees = c(150, 300),
  tree_depth = c(2, 4),
  learn_rate = c(0.03, 0.1),
  sample_size = c(0.8, 1.0),
  mtry = c(12, 24)
)

set.seed(42)
xgb_tuned <- tune_grid(
  xgb_tune_wf,
  resamples = cv_folds,
  grid = xgb_grid,
  metrics = classification_metrics,
  control = control_grid(save_pred = TRUE)
)

show_best(xgb_tuned, metric = "f_meas")
```

```{r select-best-xgboost}
best_xgb <- select_best(xgb_tuned, metric = "f_meas")
best_xgb

final_xgb_wf <- finalize_workflow(xgb_tune_wf, best_xgb)
```

## Question 7

Which parameters were selected for XGBoost? Were the selected trees shallow or deep? What does `sample_size` do?

# 13. Final Test-Set Evaluation

Now evaluate the selected models on the test set.

The test set should be used only after model comparison and tuning decisions are complete.

```{r final-test-evaluation}
final_workflows <- list(
  "Decision tree" = candidate_workflows[["Decision tree"]],
  "Random forest" = candidate_workflows[["Random forest"]],
  "Tuned XGBoost" = final_xgb_wf
)

final_predictions <- imap_dfr(
  final_workflows,
  function(wf, model_name) {
    fit(wf, data = train_data) %>%
      predict(test_data, type = "class") %>%
      bind_cols(test_data %>% select(land_cover)) %>%
      mutate(model = model_name)
  }
)

final_predictions %>%
  group_by(model) %>%
  accuracy(truth = land_cover, estimate = .pred_class)
```

```{r final-f1}
final_predictions %>%
  group_by(model) %>%
  f_meas(truth = land_cover, estimate = .pred_class)
```

```{r confusion-matrices, fig.width=12, fig.height=10}
confusion_plot_data <- final_predictions %>%
  count(model, land_cover, .pred_class, name = "n") %>%
  complete(model, land_cover, .pred_class, fill = list(n = 0))

ggplot(confusion_plot_data, aes(x = .pred_class, y = land_cover, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  facet_wrap(~ model) +
  scale_x_discrete(limits = rev) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion matrices for final test set",
    x = "Predicted class",
    y = "Observed class",
    fill = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Question 8

Which model performs best on the test set? Is that the same model that performed best during cross-validation? If not, why might that happen?

# 14. Feature Importance and Permutation Importance

Tree-based models often provide built-in feature importance, but those values can be biased. Permutation importance is usually easier to explain:

> A feature is important if model performance gets worse when that feature is randomly shuffled.

Here we will use permutation importance on the tuned XGBoost model.

```{r permutation-importance}
final_xgb_fit <- fit(final_xgb_wf, data = train_data)

baseline_predictions <- predict(final_xgb_fit, test_data, type = "class") %>%
  bind_cols(test_data %>% select(land_cover))

baseline_f1 <- f_meas(
  baseline_predictions,
  truth = land_cover,
  estimate = .pred_class
) %>%
  pull(.estimate)

set.seed(42)
permute_one_feature <- function(feature_name, repeats = 10) {
  map_dfr(seq_len(repeats), function(rep_id) {
    permuted_test <- test_data
    permuted_test[[feature_name]] <- sample(permuted_test[[feature_name]])

    permuted_predictions <- predict(final_xgb_fit, permuted_test, type = "class") %>%
      bind_cols(test_data %>% select(land_cover))

    permuted_f1 <- f_meas(
      permuted_predictions,
      truth = land_cover,
      estimate = .pred_class
    ) %>%
      pull(.estimate)

    tibble(
      feature = feature_name,
      repeat_id = rep_id,
      importance = baseline_f1 - permuted_f1
    )
  })
}

importance_df <- map_dfr(feature_cols, permute_one_feature) %>%
  group_by(feature) %>%
  summarize(
    importance_mean = mean(importance),
    importance_sd = sd(importance),
    .groups = "drop"
  ) %>%
  arrange(desc(importance_mean))

importance_df %>% slice_head(n = 12)
```

```{r permutation-importance-plot, fig.width=8, fig.height=6}
importance_df %>%
  slice_max(importance_mean, n = 12) %>%
  ggplot(aes(x = reorder(feature, importance_mean), y = importance_mean)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Permutation importance: tuned XGBoost",
    x = "Feature",
    y = "Decrease in macro F1 after permutation"
  ) +
  theme_minimal(base_size = 13)
```

## Question 9

Which spectral features appear most important? Do the central-pixel bands appear useful? Why might neighboring-pixel spectral values also help classify land cover?

Be careful: feature importance describes what the model used. It does not prove causation.

# 15. Best-Practice Reflection

Answer these questions in a few sentences each.

1. Why is a single decision tree easy to interpret but prone to overfitting?
2. Why do decision tree boundaries look blocky in two dimensions?
3. How is a random forest different from a single tree?
4. How is boosting different from random forest bagging?
5. Why should XGBoost be tuned with cross-validation instead of directly on the test set?
6. Why is permutation importance useful, and what are its limitations?
7. For environmental remote-sensing datasets, why might random train/test splits still be too optimistic?

# Key Takeaways

- A decision tree makes one-feature-at-a-time splits, creating axis-aligned boundaries.
- A shallow tree may underfit; a deep tree may overfit.
- Random forests average many trees to reduce variance.
- XGBoost builds trees sequentially so later trees correct earlier mistakes.
- XGBoost is powerful, but tuning and validation matter.
- Workflows help prevent preprocessing leakage.
- Feature importance helps explain model behavior, but importance is not causality.
- In environmental science, spatial and temporal validation may matter more than small differences among algorithms.
