---
title: "In-Class Activity 5: Logistic Regression and Support Vector Machines"
author: "GEOL 599: Machine Learning"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
# If needed, install these packages first:
# install.packages(c("tidyverse", "tidymodels", "kernlab"))

library(tidyverse)
library(tidymodels)
library(kernlab)

tidymodels_prefer()
set.seed(30)
```

# Overview

In this activity, you will compare **logistic regression** and **support vector machines (SVMs)** using the Palmer Penguins dataset.

The dataset contains field measurements from penguins observed in Antarctica. We will use two bill measurements to classify penguin species and visualize how different SVM kernels change the decision boundary.

## Learning Goals

By the end of this activity, you should be able to:

- Fit a logistic regression classifier with `tidymodels`.
- Fit SVM classifiers with linear, polynomial, and radial basis function kernels.
- Explain why feature scaling matters for SVMs.
- Compare model performance using accuracy, confusion matrices, precision, recall, and F1-score.
- Visualize how different SVM kernels change the decision boundary.

# Load the Dataset

We will use the Palmer Penguins dataset. The dataset is loaded from a CSV file.

```{r load-data}
data_url <- "https://raw.githubusercontent.com/dblaskey/ML_Course_Code/main/Data/penguins.csv"

penguins <- read_csv(data_url, show_col_types = FALSE)

glimpse(penguins)
```

# Prepare a Binary Classification Problem

To make the decision boundaries easy to visualize, we will classify only two species:

- `Chinstrap`
- `Gentoo`

We will use two predictors:

- `bill_length_mm`
- `bill_depth_mm`

This is a simplified model. In a real analysis, we might use more predictors, but using two features lets us plot the decision boundary directly.

```{r prepare-data}
species_to_keep <- c("Chinstrap", "Gentoo")

penguins_binary <- penguins %>%
  filter(species %in% species_to_keep) %>%
  select(species, bill_length_mm, bill_depth_mm) %>%
  drop_na() %>%
  mutate(species = factor(species, levels = species_to_keep))

penguins_binary %>%
  slice_head(n = 6)
```

## Question 1

How many observations are in each class? Is the dataset balanced?

**Hint:** Use `count()` on the target column.

```{r question-1}
# TODO: count the number of observations in each species class

```

# Visualize the Data

Before fitting models, plot the two predictors. This helps us anticipate whether a linear boundary might work.

```{r plot-data}
ggplot(penguins_binary, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3, alpha = 0.9) +
  labs(
    title = "Palmer Penguins: Two-Species Classification Problem",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  theme_minimal(base_size = 13)
```

## Question 2

Based on the scatterplot, do you expect a linear classifier to work well? Why or why not?

# Train/Test Split

Now split the data into training and testing sets.

Use `strata = species` so that both the training and testing sets preserve the class proportions.

```{r split-data}
# TODO: create a train/test split
# Use prop = 0.75 and strata = species

penguin_split <- initial_split(penguins_binary, prop = 0.75, strata = 'species')

train_data <- training(penguin_split)
test_data <- testing(penguin_split)

train_data %>% count(species)
test_data %>% count(species)
```

# Preprocessing Recipe

SVMs are sensitive to feature scale, so we will normalize the numeric predictors. Logistic regression also benefits from this when we want to compare coefficients or model behavior across predictors.

```{r recipe}
penguin_recipe <- recipe(species ~ bill_length_mm + bill_depth_mm, data = train_data) %>%
  step_normalize(all_numeric_predictors())

penguin_recipe
```

# Define the Models

We will compare four classifiers:

1. Logistic regression
2. Linear SVM
3. Polynomial-kernel SVM
4. RBF-kernel SVM

In R, these models are specified with `parsnip` and then combined with the preprocessing recipe using a `workflow()`.

```{r model-specs}
model_specs <- list(
  "Logistic regression" = logistic_reg() %>%
    set_mode("classification") %>%
    set_engine("glm"),

  "Linear SVM" = svm_linear(cost = 1) %>%
    set_mode("classification") %>%
    set_engine("kernlab"),

  "Polynomial SVM" = svm_poly(cost = 1, degree = 3, scale_factor = 1) %>%
    set_mode("classification") %>%
    set_engine("kernlab"),

  "RBF SVM" = svm_rbf(cost = 1, rbf_sigma = 0.1) %>%
    set_mode("classification") %>%
    set_engine("kernlab")
)
```

# Fit and Evaluate the Models

Fit each model on the training data and evaluate it on the test data.

```{r fit-models}
fit_one_model <- function(model_spec) {
  workflow() %>%
    add_recipe(penguin_recipe) %>%
    add_model(model_spec) %>%
    fit(data = train_data)
}

model_fits <- map(model_specs, fit_one_model)
```

```{r test-predictions}
test_predictions <- imap_dfr(
  model_fits,
  function(fitted_workflow, model_name) {
    predict(fitted_workflow, test_data, type = "class") %>%
      bind_cols(test_data %>% select(species)) %>%
      mutate(model = model_name)
  }
)

test_predictions %>%
  slice_head(n = 10)
```

```{r accuracy-results}
accuracy_results <- test_predictions %>%
  group_by(model) %>%
  accuracy(truth = species, estimate = .pred_class) %>%
  arrange(desc(.estimate))

accuracy_results
```

## Question 3

Which model has the highest test accuracy? Are the differences large enough to be meaningful?

# Confusion Matrices

Accuracy is useful, but it does not show which classes are being confused. Create a confusion matrix for each model.

```{r confusion-matrices}
confusion_data <- test_predictions %>%
  group_by(model) %>%
  conf_mat(truth = species, estimate = .pred_class)

confusion_data
```

```{r plot-confusion-matrices}
confusion_plot_data <- test_predictions %>%
  count(model, species, .pred_class, name = "n") %>%
  complete(model, species, .pred_class, fill = list(n = 0))

ggplot(confusion_plot_data, aes(x = .pred_class, y = species, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 5) +
  facet_wrap(~ model) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrices",
    x = "Predicted species",
    y = "Observed species",
    fill = "Count"
  ) +
  theme_minimal(base_size = 13)
```

## Question 4

Do any models make noticeably different types of mistakes? Explain using the confusion matrices.

# Classification Metrics

A classification summary gives metrics such as precision, recall, and F1-score.

```{r classification-metrics}
classification_metrics <- metric_set(accuracy, precision, recall, f_meas)

class_metric_results <- test_predictions %>%
  group_by(model) %>%
  classification_metrics(truth = species, estimate = .pred_class)

class_metric_results
```

# Visualize Decision Boundaries

The function below draws the classification regions learned by each model.

The background color shows the predicted class across the feature space. The points show the actual observations.

- Circles = training data
- X symbols = test data
- Point color = observed species
- Background color = model-predicted species

```{r decision-boundary-function}
plot_decision_boundary <- function(fitted_workflow, train_data, test_data, title) {
  x_min <- min(train_data$bill_length_mm, test_data$bill_length_mm) - 1
  x_max <- max(train_data$bill_length_mm, test_data$bill_length_mm) + 1
  y_min <- min(train_data$bill_depth_mm, test_data$bill_depth_mm) - 1
  y_max <- max(train_data$bill_depth_mm, test_data$bill_depth_mm) + 1

  prediction_grid <- expand_grid(
    bill_length_mm = seq(x_min, x_max, length.out = 250),
    bill_depth_mm = seq(y_min, y_max, length.out = 250)
  )

  grid_predictions <- predict(fitted_workflow, prediction_grid, type = "class") %>%
    bind_cols(prediction_grid)

  points_data <- bind_rows(
    train_data %>% mutate(split = "Training data"),
    test_data %>% mutate(split = "Test data")
  )

  ggplot() +
    geom_raster(
      data = grid_predictions,
      aes(x = bill_length_mm, y = bill_depth_mm, fill = .pred_class),
      alpha = 0.25
    ) +
    geom_point(
      data = points_data,
      aes(x = bill_length_mm, y = bill_depth_mm, color = species, shape = split),
      size = 3,
      stroke = 1
    ) +
    scale_fill_manual(values = c("Chinstrap" = "#56B4E9", "Gentoo" = "#E69F00")) +
    scale_color_manual(values = c("Chinstrap" = "#0072B2", "Gentoo" = "#D55E00")) +
    scale_shape_manual(values = c("Training data" = 16, "Test data" = 4)) +
    labs(
      title = title,
      x = "Bill length (mm)",
      y = "Bill depth (mm)",
      fill = "Predicted species",
      color = "Observed species",
      shape = "Dataset split"
    ) +
    coord_cartesian(xlim = c(x_min, x_max), ylim = c(y_min, y_max)) +
    theme_minimal(base_size = 12)
}
```

```{r plot-decision-boundaries, fig.width=11, fig.height=9}
decision_boundary_plots <- imap(
  model_fits,
  ~ plot_decision_boundary(.x, train_data, test_data, .y)
)

# Display the four plots together.
# If patchwork is not installed, run install.packages("patchwork") or print each plot one at a time.
if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  wrap_plots(decision_boundary_plots, ncol = 2)
} else {
  walk(decision_boundary_plots, print)
}
```

## Question 5

Compare the decision boundaries.

- Which models produce straight decision boundaries?
- Which models produce curved decision boundaries?
- Which model seems most appropriate for this dataset?

# Kernel Exploration

Now experiment with SVM settings.

Try changing:

- `cost`
- `degree` for the polynomial kernel
- `rbf_sigma` for the RBF kernel

## Question 6

Choose one SVM kernel and change one parameter. What changed in the decision boundary?

**Hint:** Smaller `cost` usually allows a wider margin with more tolerance for mistakes. Larger `cost` tries harder to classify the training data correctly. For RBF kernels, larger `rbf_sigma` can create more local, flexible decision boundaries.

```{r kernel-experiment, fig.width=7, fig.height=5}
# TODO: modify one SVM model below and visualize its decision boundary

experiment_spec <- svm_ #TODO

experiment_fit <- workflow() %>%
  add_recipe(penguin_recipe) %>%
  add_model(experiment_spec) %>%
  fit(data = train_data)

plot_decision_boundary(
  experiment_fit,
  train_data,
  test_data,
  "Experimental SVM"
)
```

```{r experiment-accuracy}
experiment_predictions <- predict(experiment_fit, test_data, type = "class") %>%
  bind_cols(test_data %>% select(species))

experiment_predictions %>%
  accuracy(truth = species, estimate = .pred_class)
```

# Reflection

Answer the following questions in a few sentences each.

1. Why do we scale predictors before fitting an SVM?
2. How is logistic regression different from a linear SVM?
3. What does the kernel do in an SVM?
4. Why might a highly flexible kernel overfit?
5. In environmental science, when might a nonlinear decision boundary be useful?
